---
title: "New Taipei Income per Household (Tax Year 2022)"
author: "User"
format: html
---

This notebook documents how the per-village average comprehensive income per household is calculated from `data/income_distribution_by_village_2022_newTaipei.csv` (see codebook: `codebook/income_distribution_by_village_2022_newTaipei.md`).

## Setup

```{r}
library(dplyr)
library(readr)
```

## Load data

CSV includes a UTF-8 BOM; read with the matching encoding and keep the English headers provided in the codebook.

```{r}
income_path <- "../data/income_distribution_by_village_2022_newTaipei.csv"
income_raw <- read.csv(income_path, check.names = FALSE, fileEncoding = "UTF-8-BOM")

dglimpse <- function(df) {
  tibble::tibble(
    n_rows = nrow(df),
    n_cols = ncol(df),
    sample_region = df$`Region Name`[1]
  )
}
dglimpse(income_raw)
```

## Calculate mean income per household

Rename columns to concise snake_case and compute `mean_income_calc = total_income / tax_units`. The dataset already reports a mean; we calculate it to verify consistency.

```{r}
income_ntpc <- income_raw %>%
  rename(
    region = `Region Name`,
    village = `Village Name`,
    tax_units = `Tax Units (Households)`,
    total_income = `Total Comprehensive Income`,
    mean_income_reported = `Mean Income`
  ) %>%
  mutate(mean_income_calc = total_income / tax_units)

income_ntpc %>%
  select(region, village, tax_units, total_income, mean_income_reported, mean_income_calc) %>%
  slice_head(n = 5)
```

## Validate reported mean vs. calculated mean

Check that the reported mean matches the calculation across all villages.

```{r}
income_ntpc %>%
  mutate(diff = mean_income_reported - mean_income_calc) %>%
  summarize(
    max_abs_diff = max(abs(diff), na.rm = TRUE),
    mean_abs_diff = mean(abs(diff), na.rm = TRUE)
  )
```

## Top and bottom villages by calculated mean income

Review extremes to understand the spread.

```{r}
# Top 5
income_ntpc %>%
  arrange(desc(mean_income_calc)) %>%
  select(region, village, tax_units, mean_income_calc, mean_income_reported) %>%
  slice_head(n = 5)

# Bottom 5
income_ntpc %>%
  arrange(mean_income_calc) %>%
  select(region, village, tax_units, mean_income_calc, mean_income_reported) %>%
  slice_head(n = 5)
```

## Notes

- Monetary unit is not stated in the CSV; based on similar MOF tables it is likely thousand NTD—confirm with source metadata.
- If differences appear between reported and calculated means, revisit unit assumptions and rounding rules from the source.
